---
title: "Lab 07b: PCA Example: MNIST Handwritten Digits"
subtitle: "STT843 Multivariate Data Analysis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
# Note: Ensure the 'dslabs' package is installed for the MNIST data
# install.packages("dslabs")
library(dslabs)
```

## MNIST: The 784-Dimensional Space

Handwritten digits in the MNIST dataset are $28 \times 28$ pixel images.

Each pixel is a feature (variable).

$p = 28 \times 28 = 784$.

Each image is a vector $\mathbf{x}_i \in \mathbb{R}^{784}$.

PCA finds the best $k$-dimensional subspace to represent these 784
dimensions while minimizing "ink" loss (reconstruction error).

## Loading the Data 
We will look at a subset of the digits (the "zeros")

```{r}
mnist <- read_mnist()
# Filter for the digit '0'
zeros <- mnist$train$images[mnist$train$labels == 0, ]
n_images <- nrow(zeros)

# Center the data
zeros_centered <- scale(zeros, center = TRUE, scale = FALSE)
avg_zero <- attr(zeros_centered, "scaled:center")
```

## Visualizing the "Average Zero" 
The "center" of our data
$\bar{\mathbf{x}}$ is the average shape of a handwritten zero.

```{r}
image(matrix(avg_zero, 28, 28)[, 28:1], col = gray.colors(256), 
      main = "The Mean Digit (Origin)", xaxt='n', yaxt='n')
      
 # Get the indices of the first 9 occurrences of the digit '0'
zero_indices <- which(mnist$train$labels == 0)[1:9]

par(mfrow = c(3, 3), mar = c(0.1, 0.1, 1, 0.1))

for (idx in zero_indices) {
  mat <- matrix(mnist$train$images[idx, ], 28, 28)
  image(mat[, 28:1], col = gray.colors(256), xaxt = 'n', yaxt = 'n')
}
```

## Eigen-Digits (Principal Components)

The eigenvectors $\mathbf{v}_j$ (Loadings) represent the primary modes
of variation. In image analysis, these are often called Eigen-digits.

```{r}
# We use svd for efficiency on large matrices
pca_zeros <- prcomp(zeros, center = TRUE, scale. = FALSE)

# Variance Explained
var_exp <- cumsum(pca_zeros$sdev^2) / sum(pca_zeros$sdev^2)
```

PC1 typically captures the "thickness" or "tilt" of the zero.

## The Scree Plot: How many dimensions? 
How many components do we need to
reconstruct a '0' accurately?

```{r}
plot(var_exp[1:100], type = "b", xlab = "Number of Components", 
     ylab = "Cumulative Variance Explained", main = "Scree Plot for MNIST '0's")
abline(h = 0.9, col = "red", linetype = "dashed")
```

## Reconstruction: PCA as Least Squares Recall
$\hat{\mathbf{x}}_i = \bar{\mathbf{x}} + \sum_{j=1}^k y_{ij}\mathbf{v}_j$.As
$k$ increases, the approximation $\hat{\mathbf{x}}_i$ gets closer to the
original $\mathbf{x}_i$.

```{r}
# Reconstruct the first '0' using k components
reconstruct <- function(pca_obj, k, index, mean_vec) {
  recon <- mean_vec + pca_obj$x[index, 1:k] %*% t(pca_obj$rotation[, 1:k])
  return(matrix(recon, 28, 28)[, 28:1])
}

par(mfrow=c(1,4))
image(reconstruct(pca_zeros, 2, 1, avg_zero), col=gray.colors(256), main="k=2", xaxt='n', yaxt='n')
image(reconstruct(pca_zeros, 10, 1, avg_zero), col=gray.colors(256), main="k=10", xaxt='n', yaxt='n')
image(reconstruct(pca_zeros, 50, 1, avg_zero), col=gray.colors(256), main="k=50", xaxt='n', yaxt='n')
image(matrix(zeros[1,], 28, 28)[, 28:1], col=gray.colors(256), main="Original", xaxt='n', yaxt='n')
```
