---
title: "Lab08a: Factor Analysis and Structured Covariance Matrices"
subtitle: "STT 843: Multivariate Analysis"
author: "Dr. Guanqun Cao"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(psych)  # For factor analysis functions
library(GPArotation) # For Varimax and other rotations
```

##1. Introduction to the Orthogonal Factor Model

The purpose of factor analysis is to describe the covariance relationships among $p$ variables in terms of a few underlying, unobservable random quantities called **factors**.

The model is:
$$\boldsymbol{X}-\boldsymbol{\mu}=\mathbf{L F}+\boldsymbol\varepsilon$$

Where:
- $\mathbf{L}$ is the matrix of *factor loadings*.
- $\mathbf{F}$ are the *common factors*.
- $\boldsymbol\varepsilon$ are the *specific factors* (errors).

### Covariance Structure
Under the model assumptions ($Cov(F)=I$, $Cov(\varepsilon)=\Psi$):
$$\operatorname{Cov}(\boldsymbol{X}) = \mathbf{L L}^{\prime} + \boldsymbol{\Psi}$$

---

## 2. Theoretical Example: Verifying Factorability

Consider the covariance matrix $\boldsymbol{\Sigma}$:
```{r}
Sigma <- matrix(c(19, 30, 2, 12,
                  30, 57, 5, 23,
                  2, 5, 38, 47,
                  12, 23, 47, 68), nrow = 4, byrow = TRUE)
colnames(Sigma) <- rownames(Sigma) <- c("X1", "X2", "X3", "X4")
Sigma
```

If we postulate two factors, we check if $\boldsymbol{\Sigma} \approx \mathbf{LL}' + \boldsymbol{\Psi}$.

---

## 3. Principal Component Solution

In this method, we estimate loadings using eigenvalue-eigenvector pairs of $\mathbf{S}$ or $\mathbf{R}$.

### Example: Consumer-Preference Data
We use the correlation matrix provided in the notes.

```{r}
# Constructing the correlation matrix
R_consumer <- matrix(c(1.00, 0.02, 0.96, 0.42, 0.01,
                       0.02, 1.00, 0.13, 0.71, 0.85,
                       0.96, 0.13, 1.00, 0.50, 0.11,
                       0.42, 0.71, 0.50, 1.00, 0.79,
                       0.01, 0.85, 0.11, 0.79, 1.00), nrow = 5)
colnames(R_consumer) <- c("Taste", "Money", "Flavor", "Snack", "Energy")

# Principal Component Method (nfactors = 2)
pc_fit <- principal(R_consumer, nfactors = 2, rotate = "none")
print(pc_fit)
```

---

## 4. Factor Rotation (Varimax)

Loadings are determined only up to an orthogonal matrix $\mathbf{T}$. *Varimax* rotation seeks to maximize the variance of the squared loadings to simplify interpretation.

```{r}
# Applying Varimax Rotation to the consumer data
rotated_fit <- principal(R_consumer, nfactors = 2, rotate = "varimax")
print(rotated_fit$loadings, cutoff = 0)
```

---

## 5. Maximum Likelihood Method

The ML method assumes multivariate normality and maximizes the likelihood function.

### Example: Stock-Price Data
Using the `factanal()` function (which defaults to Maximum Likelihood).

```{r}
# Example using built-in dataset or reconstructed R for stocks
# factanal requires raw data or a covariance matrix + n
# Here we demonstrate the function call structure:
# ml_fit <- factanal(factors = 2, covmat = R_stocks, n.obs = 103, rotation = "varimax")
```

 

---

## 6. Computing Factor Scores

Factor scores are estimates of the values of $\mathbf{F}$ for each observation.

### Weighted Least Squares (Bartlett) vs. Regression Method
In R, we specify the `scores` argument in `factanal`.

```{r, eval=FALSE}
# Regression method
fit_reg <- factanal(my_data, factors = 2, scores = "regression")

# Bartlett (Weighted Least Squares) method
fit_bart <- factanal(my_data, factors = 2, scores = "Bartlett")
```

 
### Key R Functions Used:
* **`principal()`**: From the `psych` package; used for the Principal Component solution of the factor model.
* **`factanal()`**: Built-in R function; performs Maximum Likelihood Factor Analysis.
* **`rotate = "varimax"`**: Argument used to apply the Varimax criterion for simple structure.
* **`scores = "regression"`**: Argument used to calculate factor scores based on the regression approach.