---
title: "Untitled"
output: html_document
date: "2026-01-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(MASS)       # For mvrnorm
library(matrixcalc) # For matrix properties
```

## 

Here is a complete R Markdown lab based on your lecture notes. This lab is designed to help students practically explore the properties of the Multivariate Normal Distribution (MND) and Maximum Likelihood Estimation (MLE) using R.

------------------------------------------------------------------------

### R Markdown File Content

Markdown

```         
---
title: "Lab: Multivariate Normal Distribution and MLE"
subtitle: "Multivariate Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: united
---

```{r setup, include=T}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(MASS)       # For mvrnorm
library(matrixcalc) # For matrix properties
```

## 1. Introduction to the Multivariate Normal Distribution (MND)

In the lecture, we discussed that if $\boldsymbol{X} \sim N_p(\boldsymbol{\mu}, \Sigma)$, then any linear combination $\mathbf{A}\boldsymbol{X}$ is also normally distributed.

### Exercise 1: Linear Combinations (Example 3.4)

Let \$\\boldsymbol{X} \\sim N_3(\\boldsymbol{\\mu}, \\Sigma)\$ where:

\$\$\\boldsymbol{\\mu} = \\begin{bmatrix} 1 \\\\ 2 \\\\ 3 \\end{bmatrix}, \\Sigma = \\begin{bmatrix} 4 & 1 & 0 \\\\ 1 & 3 & 0 \\\\ 0 & 0 & 2 \\end{bmatrix}\$\$

We want to find the distribution of $\mathbf{A}\boldsymbol{X}$ where $\mathbf{A} = \begin{bmatrix} 1 & -1 & 0 \\ 0 & 1 & -1 \end{bmatrix}$.

```{r setup, include=T}
# Define parameters
mu <- c(1, 2, 3)
Sigma <- matrix(c(4, 1, 0, 
                  1, 3, 0, 
                  0, 0, 2), nrow = 3, byrow = TRUE)
A <- matrix(c(1, -1, 0, 
              0, 1, -1), nrow = 2, byrow = TRUE)

# Theoretical New Mean: A * mu
new_mu <- A %*% mu

# Theoretical New Covariance: A * Sigma * A^T
new_Sigma <- A %*% Sigma %*% t(A)

print("New Mean Vector:")
print(new_mu)
print("New Covariance Matrix:")
print(new_Sigma)
```
