---
title: "Lab09a: Canonical Correlation Analysis: Foundations and Application"
subtitle: "STT 843: Multivariate Analysis"
author: "Dr. Guanqun Cao"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---
 
 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(CCA)
```

# 1. Introduction to Canonical Correlation
Canonical Correlation Analysis (CCA) is used to identify and quantify the associations between two sets of variables. It finds linear combinations of variables in the first set ($X^{(1)}$) and the second set ($X^{(2)}$) that have the highest possible correlation.

**Examples:**
* **Education:** Arithmetic skills vs. Reading skills.
* **Economics:** Government policy variables vs. Economic goal indicators.
* **Psychology:** Job characteristics vs. Job satisfaction.

# 2. Mathematical Formulation
Let $\boldsymbol{X}^{(1)}$ be $p \times 1$ and $\boldsymbol{X}^{(2)}$ be $q \times 1$ ($p \leq q$).
We seek vectors $\mathbf{a}$ and $\mathbf{b}$ to maximize:
$$\operatorname{Corr}(U, V) = \frac{\mathbf{a}^{\prime} \Sigma_{12} \mathbf{b}}{\sqrt{\mathbf{a}^{\prime} \Sigma_{11} \mathbf{a}} \sqrt{\mathbf{b}^{\prime} \Sigma_{22} \mathbf{b}}}$$



### Result 9.1: Canonical Variate Pairs
The $k$-th pair of canonical variates $(U_k, V_k)$ are linear combinations that maximize correlation subject to being uncorrelated with the previous $k-1$ pairs.

* **First Pair ($U_1, V_1$):** Highest possible correlation $\rho_1^*$.
* **Properties:** $\operatorname{Var}(U_k) = \operatorname{Var}(V_k) = 1$ and $\operatorname{Corr}(U_k, U_l) = \operatorname{Corr}(V_k, V_l) = 0$ for $k \neq l$.

# 3. Sample Canonical Correlation
In practice, we use the sample covariance matrices $S_{11}, S_{22}, S_{12}$ to estimate the variates.

### Calculation Logic
The squared sample canonical correlations $\hat{\rho}_k^{*2}$ are the eigenvalues of:
$$S_{11}^{-1/2} S_{12} S_{22}^{-1} S_{21} S_{11}^{-1/2}$$

The sample canonical variates are then:
$$\hat{U}_k = \hat{\mathbf{a}}_k^{\prime} \mathbf{x}^{(1)}, \quad \hat{V}_k = \hat{\mathbf{b}}_k^{\prime} \mathbf{x}^{(2)}$$

# 4. Example: Chicken-Bone Data (Standardized)
Analysis of skull measurements vs. leg measurements in white leghorn fowl.

```{r chicken_bone}
# Correlation Matrix R
R <- matrix(c(1.0, .505, .569, .602,
              .505, 1.0, .422, .467,
              .569, .422, 1.0, .926,
              .602, .467, .926, 1.0), 4, 4)

# Canonical Correlation Result (First Pair)
rho1 <- 0.631
# U1 = 0.781*z1 + 0.345*z2
# V1 = 0.060*z3 + 0.944*z4
```

**Interpretation:** The first canonical correlation is moderately strong (0.631). $U_1$ is heavily influenced by skull length, while $V_1$ is almost entirely dominated by tibia length.

# 5. Case Study: Job Satisfaction
In a study of $n=784$ executives, we analyze $p=5$ job characteristics and $q=7$ satisfaction variables.

### Interpreting Correlations (Loadings)
It is often more useful to look at the **loadings** (correlations between the canonical variate and the original variables) rather than just the coefficients.

| Variable | Corr with $\hat{U}_1$ | Corr with $\hat{V}_1$ |
| :--- | :---: | :---: |
| Feedback | 0.83 | 0.46 |
| Autonomy | 0.85 | 0.48 |
| Supervisor Sat. | 0.42 | 0.75 |
| Kind-of-work Sat. | 0.44 | 0.80 |

**Conclusion:** The first pair ($\hat{\rho}_1^* = 0.55$) suggests a "Job Characteristic Index" ($U_1$) is related to a "Satisfaction-Identification Index" ($V_1$).

# 6. Best Practices for Analysis
* **Scatter Plots:** Always plot $(\hat{U}_k, \hat{V}_k)$ to identify outliers or non-linear patterns.
* **Sample Splitting:** For large datasets (like the job satisfaction study), split the data to validate the stability of the canonical coefficients.
* **Standardization:** Standardization is recommended when variables are measured in different units; it does not change the canonical correlations themselves.
```