---
title: "Lab 10b: Discrimination and Classification"
subtitle: "Multivariate Statistical Analysis"
author: "Guanqun Cao"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

#  Setup and Data Preparation

First, we will recreate the Riding Mower dataset. This is a classic "Two Populations" problem with two predictors: Income ($x_1$) and Lot Size ($x_2$).

```{r setup}
# Load necessary libraries

library(MASS)    # For lda() and qda()
library(ggplot2) # For visualization

# Create the dataset from lecture notes

mowers <- data.frame(
  Income = c(90.0, 115.5, 94.8, 91.5, 117.0, 140.1, 138.0, 112.8, 99.0, 123.0, 81.0, 111.0,
             105.0, 82.8, 94.8, 73.2, 114.0, 79.2, 89.4, 96.0, 77.4, 63.0, 81.0, 93.0),
  LotSize = c(18.4, 16.8, 21.6, 20.8, 23.6, 19.2, 17.6, 22.4, 20.0, 20.8, 22.0, 20.0,
              19.6, 20.8, 17.2, 20.4, 17.6, 17.6, 16.0, 18.4, 16.4, 18.8, 14.0, 14.8),
  Owner = factor(c(rep("Owner", 12), rep("Nonowner", 12)))
)

# Visualizing the separation (Goal 1: Graphical Description)
ggplot(mowers, aes(x=Income, y=LotSize, color=Owner, shape=Owner)) +
  geom_point(size=3) +
  theme_minimal() +
  labs(title = "Riding Mower Owners vs. Nonowners")
```
#2. Linear Discriminant Analysis (LDA)

LDA assumes $\Sigma_1 = \Sigma_2 = \Sigma$ (equal covariance matrices) and multivariate normality.

```{r}
# Fit the LDA model
lda_model <- lda(Owner ~ Income + LotSize, data = mowers)


# Display the coefficients (Fisher's linear combination)
print(lda_model)


# Calculate the Midpoint/Decision Boundary
# Note: lda() in R uses a slightly different scaling than the â vector 
# in the notes, but the decision rule remains equivalent.
```
## 3. Evaluating Performance: APER and Confusion Matrix
We use the Confusion Matrix to calculate the Apparent Error Rate (APER).

```{r}
# Predictions on the training set
lda_pred <- predict(lda_model, mowers)
conf_matrix <- table(Actual = mowers$Owner, Predicted = lda_pred$class)

print(conf_matrix)
 
# Calculate APER
 
aper <- (conf_matrix[1,2] + conf_matrix[2,1]) / sum(conf_matrix)
cat("Apparent Error Rate (APER):", round(aper * 100, 2), "%\n")
```

4. Lachenbruch’s Holdout Procedure (Leave-One-Out CV)

The lda() function has a built-in parameter CV = TRUE to perform the "holdout" procedure described in your notes.

```{r}
# Perform Leave-One-Out Cross-Validation
lda_cv <- lda(Owner ~ Income + LotSize, data = mowers, CV = TRUE)

# Confusion Matrix for Holdout Procedure
cv_matrix <- table(Actual = mowers$Owner, Predicted = lda_cv$class)
print(cv_matrix)

# Estimated Actual Error Rate (AER)
aer_hat <- (cv_matrix[1,2] + cv_matrix[2,1]) / sum(cv_matrix)
cat("Estimated AER (Holdout):", round(aer_hat * 100, 2), "%\n")
```

5. Quadratic Discriminant Analysis (QDA)

When the assumption $\Sigma_1 = \Sigma_2$ is violated, we use QDA.


```{r}
# Fit QDA model
qda_model <- qda(Owner ~ Income + LotSize, data = mowers)

# Predict for a new observation (e.g., Income=100, LotSize=19)
 
new_obs <- data.frame(Income = 100, LotSize = 19)
predict(qda_model, new_obs)
```

6. Fisher’s Method (Multi-class Example)
For more than two populations, we can visualize the separation using the first two discriminants.
```{r}
# Using the built-in Iris dataset for multiple populations (3 species)

data(iris)
iris_lda <- lda(Species ~ ., data = iris)

# Projecting data onto the first two discriminants (LD1 and LD2)
```{r}
lda_values <- predict(iris_lda)
plot_data <- data.frame(lda_values$x, Species = iris$Species)

ggplot(plot_data, aes(x=LD1, y=LD2, color=Species)) +
  geom_point() +
  theme_classic() +
  labs(title = "Fisher's Discriminant Projection (Iris Data)")
```